Internal models of type theory

### 2013-10-29 article-models-tt.tex
compatibility rules and universal property of extension

\subsection{Substitution}
\begin{infarray}{cc}
\inference{\jtermt{\Gamma}{A}{x}\qquad\jtype{\ctxext{\Gamma}{A}}{P}}{\jtype{\Gamma}{\subst{x}{P}}}
& \inference{\jtermteq{\Gamma}{A}{x}{x'}\qquad\jtypeeq{\ctxext{\Gamma}{A}}{P}{P'}}{\jtypeeq{\Gamma}{\subst{x}{P}}{\subst{x'}{P'}}}\\
\inference{\jtermt{\Gamma}{A}{x}\qquad\jtype{\ctxext{{\Gamma}{A}}{P}}{Q}}{\jtype{\ctxext{\Gamma}{\subst{x}{P}}}{\subst{x}{Q}}}
& \inference{\jtermteq{\Gamma}{A}{x}{x'}\qquad\jtypeeq{\ctxext{{\Gamma}{A}}{P}}{Q}{Q'}}{\jtypeeq{\ctxext{\Gamma}{\subst{x}{P}}}{\subst{x}{Q}}{\subst{x'}{Q'}}}\\
\inference{\jtermt{\Gamma}{A}{x}\qquad\jtermt{\ctxext{\Gamma}{A}}{P}{f}}{\jtermt{\Gamma}{\subst{x}{P}}{\subst{x}{f}}}
& \inference{\jtermteq{\Gamma}{A}{x}{x'}\qquad\jtermteq{\ctxext{\Gamma}{A}}{P}{f}{f'}}{\jtermteq{\Gamma}{\subst{x}{P}}{\subst{x}{f}}{\subst{x'}{f'}}}\\
\inference{\jtermt{\Gamma}{A}{x}\qquad\jtermt{\ctxext{{\Gamma}{A}}{P}}{Q}{g}}{\jtermt{\ctxext{\Gamma}{\subst{x}{P}}}{\subst{x}{Q}}{\subst{x}{g}}}
& \inference{\jtermteq{\Gamma}{A}{x}{x'}\qquad\jtermteq{\ctxext{{\Gamma}{A}}{P}}{Q}{g}{g'}}{\jtermteq{\ctxext{\Gamma}{\subst{x}{P}}}{\subst{x}{Q}}{\subst{x}{g}}{\subst{x'}{g'}}}
\end{infarray}

\subsection{Definitional equality rules for weakening and substitution}
The judgmental equalities we're about to describe assert that weakening followed
by substitution leaves everything untouched. Thus, we get that each fiber
$\subst{x}{\ctxwk{A}{B}}$ is just $B$, that $\ctxwk{A}{y}$ is the constant function
mapping everything to $y:B$ and similar properties for families and terms thereof.

\begin{infarray}{c}
\inference{\jtype{\Gamma}{A}\qquad\jtype{\Gamma}{B}\qquad\jtermt{\Gamma}{A}{x}}{\jtypeeq{\Gamma}{\subst{x}{\ctxwk{A}{B}}}{B}}\\
\inference{\jtype{\Gamma}{A}\qquad\jtype{\ctxext{\Gamma}{B}}{Q}\qquad\jtermt{A}{x}}{\jtypeeq{\ctxext{\Gamma}{B}}{\subst{x}{\ctxwk{A}{Q}}}{Q}}\\
\inference{\jtermt{\Gamma}{A}{x}\qquad\jtermt{\Gamma}{B}{y}}{\jtermteq{\Gamma}{B}{\subst{x}{\ctxwk{A}{y}}}{y}}\\
\inference{\jtermt{\Gamma}{A}{x}\qquad\jtermt{\ctxext{\Gamma}{B}}{Q}{g}}{\jtermteq{\ctxext{\Gamma}{B}}{Q}{\subst{x}{\ctxwk{A}{g}}}{g}}
\end{infarray}

>>>>

\subsubsection{Weakening is compatible with itself}
Suppose $\jtype{\ctxext{\Gamma}{A}}{P}$ and $\jtype{\Gamma}{B}$. Then we can
consider the weakening $\ctxwk{A}{B}$ over $A$ and consecutively the weakening
$\ctxwk{P}{{A}{B}}$. This should be isomorphic to $\ctxwk{\ctxext{A}{P}}{B}$.

\subsection{The universal property of extension}

Using weakening and substitution we are able to state the universal property
for extension. It looks a bit more involved, since we cannot directly refer
to the variables in the contexts. On the other hand, we can now plainly see
were there were secretly weakenings going on.

In all of the following inference rules we assume that $\jtype{\ctxext{\Gamma}{A}}{P}$
is among the hypothesis. The induction principle for extension consists of three
inference rules:

\begin{infarray}{c}
\inference{}
{\jtermt{\ctxext{{\Gamma}{A}}{P}}{\ctxwk{P}{{A}{\ctxext{A}{P}}}}{\pair^P}}\\
\inference{
  \jtype{\ctxext{\Gamma}{{A}{P}}}{Q}
  \qquad
  \jtermt{\ctxext{{\Gamma}{A}}{P}}{\subst{\pair^P}{\ctxwk{P}{{A}{Q}}}}{f}}
  {\jtermt{\ctxext{\Gamma}{{A}{P}}}{Q}{\ind{ext}(f)}}\\
\inference{
  \jtype{\ctxext{\Gamma}{{A}{P}}}{Q}
  \qquad
  \jtermt{\ctxext{{\Gamma}{A}}{P}}{\subst{\pair^P}{\ctxwk{P}{{A}{Q}}}}{f}}
  {\jtermteq{\ctxext{{\Gamma}{A}}{P}}{\subst{\pair^P}{\ctxwk{P}{{A}{Q}}}}{\subst{\pair^P}{\ctxwk{P}{{A}{\ind{ext}(f)}}}}{f}}
\end{infarray}

\subsubsection{Weakening is compatible with extension}
Suppose $\jtype{\Gamma}{A}$ and $\jtype{\ctxext{\Gamma}{B}}{Q}$. Then
\begin{equation*}
s:\ctxwk{{A}{\ctxext{B}{Q}}}{\ctxext{\ctxwk{A}{B}}{\ctxwk{A}{Q}}}
\end{equation*}

\subsection{Substitution}
Given a family $P$ over $A$ and a term $x$ of $A$, substitution gives a way to
consider the fiber $\subst{x}{P}$ of $P$ at $x$. Also, we get a way to evaluate
terms $f$ of $P$ at $x$. This will give us ways to compose functions too. In
this section, we shall first introduce the operations `substitution of a term $x$'
for types, terms and families. Then we shall explain how substitution interacts
with itself, extension and weakening.

In the rules introducing the various substitutions we assume $\jtermt{\Gamma}{A}{x}$;
in the rules introducing the definitional equalities we assume $\jtermteq{\Gamma}{A}{x}{x'}$.

\begin{infarray}{cc}
\inference{\jtype{\ctxext{\Gamma}{A}}{P}}{\jtype{\Gamma}{\subst{x}{P}}}
& \inference{\jtypeeq{\ctxext{\Gamma}{A}}{P}{P'}}{\jtypeeq{\Gamma}{\subst{x}{P}}{\subst{x'}{P'}}}\\
\inference{\jtype{\ctxext{{\Gamma}{A}}{P}}{Q}}{\jtype{\ctxext{\Gamma}{\subst{x}{P}}}{\subst{x}{Q}}}
& \inference{\jtypeeq{\ctxext{{\Gamma}{A}}{P}}{Q}{Q'}}{\jtypeeq{\ctxext{\Gamma}{\subst{x}{P}}}{\subst{x}{Q}}{\subst{x'}{Q'}}}\\
\inference{\jtermt{\ctxext{\Gamma}{A}}{P}{f}}{\jtermt{\Gamma}{\subst{x}{P}}{\subst{x}{f}}}
& \inference{\jtermteq{\ctxext{\Gamma}{A}}{P}{f}{f'}}{\jtermteq{\Gamma}{\subst{x}{P}}{\subst{x}{f}}{\subst{x'}{f'}}}\\
\inference{\jtermt{\ctxext{{\Gamma}{A}}{P}}{Q}{g}}{\jtermt{\ctxext{\Gamma}{\subst{x}{P}}}{\subst{x}{Q}}{\subst{x}{g}}}
& \inference{\jtermteq{\ctxext{{\Gamma}{A}}{P}}{Q}{g}{g'}}{\jtermteq{\ctxext{\Gamma}{\subst{x}{P}}}{\subst{x}{Q}}{\subst{x}{g}}{\subst{x'}{g'}}}
\end{infarray}

\subsubsection{Substitution is compatible with substitution}

We require that substitution is compatible with itself, which is roughly the
assertion that substitution is associative. However, we cannot just state that
$\subst{x}{{f}{g}}\jdeq\subst{{x}{f}}{g}$ since the expression $\subst{{x}{f}}{g}$
is not well-formed. The term $\subst{x}{f}$ can be substituted in (terms of) families over
$\subst{x}{P}$; the term $\subst{x}{g}$ is such. Therefore, associativity of
substitution takes the form $\subst{x}{{f}{g}}\jdeq\subst{{x}{f}}{{x}{g}}$.

In the following inference rules we assume
$\jtermt{\Gamma}{A}{x}$ and $\jtermt{\ctxext{\Gamma}{A}}{P}{f}$.

\begin{infarray}{c}
\inference{\jtype{\ctxext{{\Gamma}{A}}{P}}{Q}}
{\jtypeeq{\Gamma}{\subst{x}{{f}{Q}}}{\subst{{x}{f}}{{x}{Q}}}}\\
\inference{\jtype{\ctxext{{{\Gamma}{A}}{P}}{Q}}{R}}
{\jtypeeq{\ctxext{\Gamma}{\subst{x}{{f}{Q}}}}{\subst{x}{{f}{R}}}{\subst{{x}{f}}{{x}{R}}}}\\
\inference{\jtermt{\ctxext{{\Gamma}{A}}{P}}{Q}{g}}
{\jtermteq{\Gamma}{\subst{x}{{f}{Q}}}{\subst{x}{{f}{g}}}{\subst{{x}{f}}{{x}{g}}}}\\
\inference{\jtermt{\ctxext{{{\Gamma}{A}}{P}}{Q}}{R}{h}}
{\jtermteq{\ctxext{\Gamma}{\subst{x}{{f}{Q}}}}{\subst{x}{{f}{R}}}{\subst{x}{{f}{h}}}{\subst{{x}{f}}{{x}{h}}}}
\end{infarray}

\subsubsection{Substitution is compatible with extension}
Suppose $\jtermt{\Gamma}{A}{x}$ in all of the following inference rules.
\begin{infarray}{c}
\inference{\jtype{\ctxext{{\Gamma}{A}}{P}}{Q}}{\jtypeeq{\Gamma}{\subst{x}{\ctxext{P}{Q}}}{\ctxext{\subst{x}{P}}{\subst{x}{Q}}}}
\end{infarray}

\subsubsection{Substitution is compatible with weakening}
The judgmental equalities we're about to describe assert that weakening followed
by substitution leaves everything untouched. Thus, we get that each fiber
$\subst{x}{\ctxwk{A}{B}}$ is just $B$, that $\ctxwk{A}{y}$ is the constant function
mapping everything to $y:B$ and similar properties for families and terms thereof.

\begin{infarray}{c}
\inference{\jtype{\Gamma}{A}\qquad\jtype{\Gamma}{B}\qquad\jtermt{\Gamma}{A}{x}}{\jtypeeq{\Gamma}{\subst{x}{\ctxwk{A}{B}}}{B}}\\
\inference{\jtype{\Gamma}{A}\qquad\jtype{\ctxext{\Gamma}{B}}{Q}\qquad\jtermt{\Gamma}{A}{x}}{\jtypeeq{\ctxext{\Gamma}{B}}{\subst{x}{\ctxwk{A}{Q}}}{Q}}\\
\inference{\jtermt{\Gamma}{A}{x}\qquad\jtermt{\Gamma}{B}{y}}{\jtermteq{\Gamma}{B}{\subst{x}{\ctxwk{A}{y}}}{y}}\\
\inference{\jtermt{\Gamma}{A}{x}\qquad\jtermt{\ctxext{\Gamma}{B}}{Q}{g}}{\jtermteq{\ctxext{\Gamma}{B}}{Q}{\subst{x}{\ctxwk{A}{g}}}{g}}
\end{infarray}
